---
title: "X-modelling — surveillance data eyeballing"
date: "`r format(Sys.Date(), '%d %B %Y')`"
output:
  html_document:
    toc: true
    toc_float: true
    theme: flatly
---

```{r setup, include=FALSE}
# Large figures throughout; each figure prints full width with its findings underneath.
knitr::opts_chunk$set(echo=FALSE, message=FALSE, warning=FALSE,
                      out.width="100%", dpi=110, fig.align="center")
library(here)

# Build the figure manifest once. If `eb` already exists in the knitting environment
# (e.g. you ran the pipeline in 00_main.R and knit from there) we reuse it; otherwise we
# rebuild from the local snapshots so the report is self-contained.
if (!exists("eb")) {
  source(here("code/01_main_supporting/setup.R"))
  source(here("code/02_settings/settings_version0.R")); params = settings()
  source(here("code/01_main_supporting/flu_functions.R"))
  source(here("code/01_main_supporting/load_data.R"))
  source(here("code/01_main_supporting/gen_model_input.R"))
  source(here("code/01_main_supporting/eyeballing.R"))
  data      = load_data(params, regenerate=FALSE, new_from_online=FALSE)
  data      = load_data_epi(data, params, regenerate=TRUE, new_from_online=FALSE) # ensure fresh epi from snapshot
  models_in = gen_model_input(params, data)
  eb        = eyeballing(models_in, params, data)
}

# small helper: print a figure's subtitle + narrative bullets as markdown
print_findings = function(fig){
  cat("*", fig$subtitle, "*\n\n", sep="")
  cat(paste0("- ", fig$bullets), sep="\n")
  cat("\n")
}
```

This report eyeballs the canonical tables built by `gen_model_input()` for
**`r eb$meta$n_countries` countries** across **`r eb$meta$n_seasons` seasons**
(`r min(eb$meta$seasons)` to `r max(eb$meta$seasons)`). It is split into
**data-quality** figures (how complete/reliable each indicator is) and
**temporal-dynamics** figures (how indicators move over time by disease and country).
Bullets are editable narrative placeholders — replace them with what you actually see.

# Data quality

## `r eb$figures$completeness$title`

```{r completeness, fig.width=15, fig.height=10}
eb$figures$completeness$plot
```

```{r completeness-bullets, results="asis"}
print_findings(eb$figures$completeness)
```

## `r eb$figures$testing_volume$title`

```{r testing-volume, fig.width=15, fig.height=9}
eb$figures$testing_volume$plot
```

```{r testing-volume-bullets, results="asis"}
print_findings(eb$figures$testing_volume)
```

# Temporal dynamics

## `r eb$figures$iliplus_dynamics$title`

```{r iliplus, fig.width=18, fig.height=12}
eb$figures$iliplus_dynamics$plot
```

```{r iliplus-bullets, results="asis"}
print_findings(eb$figures$iliplus_dynamics)
```

## `r eb$figures$iliplus_aligned$title`

```{r iliplus-aligned, fig.width=15, fig.height=22}
eb$figures$iliplus_aligned$plot
```

```{r iliplus-aligned-bullets, results="asis"}
print_findings(eb$figures$iliplus_aligned)
```

## `r eb$figures$positivity_dynamics$title`

```{r positivity, fig.width=18, fig.height=12}
eb$figures$positivity_dynamics$plot
```

```{r positivity-bullets, results="asis"}
print_findings(eb$figures$positivity_dynamics)
```

## `r eb$figures$syndromic_dynamics$title`

```{r syndromic, fig.width=18, fig.height=12}
eb$figures$syndromic_dynamics$plot
```

```{r syndromic-bullets, results="asis"}
print_findings(eb$figures$syndromic_dynamics)
```

# Supplement

Additional views worth keeping handy: an alternative completeness measure that ignores the
off-season, a cross-source ILI+ consistency check, age structure, and raw detection counts.

## `r eb$figures$completeness_active$title`

```{r completeness-active, fig.width=15, fig.height=10}
eb$figures$completeness_active$plot
```

```{r completeness-active-bullets, results="asis"}
print_findings(eb$figures$completeness_active)
```

## `r eb$figures$iliplus_source_compare$title`

```{r iliplus-source, fig.width=18, fig.height=12}
eb$figures$iliplus_source_compare$plot
```

```{r iliplus-source-bullets, results="asis"}
print_findings(eb$figures$iliplus_source_compare)
```

## `r eb$figures$age_dynamics$title`

```{r age, fig.width=18, fig.height=12}
eb$figures$age_dynamics$plot
```

```{r age-bullets, results="asis"}
print_findings(eb$figures$age_dynamics)
```

## `r eb$figures$detections_dynamics$title`

```{r detections, fig.width=18, fig.height=12}
eb$figures$detections_dynamics$plot
```

```{r detections-bullets, results="asis"}
print_findings(eb$figures$detections_dynamics)
```
